### 2024-03-05 tex/lnd.tex
[ writing ] ready for submission
[ correction ] first pass over chapter 4

The first two connectives are those we've already seen for intuitionistic
systems --- $\dot1$ and $\dottimes$.
The absence of premises is encoded by $\dot1$, while the space between premises
sharing a context is encoded by $\dottimes$.
As for implication, I temporarily avoid giving a $\dotto$ connectives, instead
fusing it together with $\forallb{-}$ to produce the \emph{set} of
context-preserving functions $T \rightarrowtriangle U$.
When we interpret a typing rule as a constructor of an inductive definition,
$\rightarrowtriangle$
interprets the horizontal line, reflecting the fact that the usage annotations
we start off with in the premises are those of the conclusion, corresponding to
a general principle of resource conservation.
The prototypical rules that use $\dot1$ and $\dottimes$ are the introduction
rules for $\top$ and $\with$, respectively.

@@

@@

The separating conjunction $\sep$ can be seen as a decategorified version of Day
convolution~\citep{Day70}.
It also resembles the use of ternary frames in semantics of non-distributive
logics~\citep[chapter 12]{Restall99}.

@@

For all uses of bunched connectives in this thesis, the carrier set $A$ will
form a partial order --- for example, with contexts, the order is given by the
pointwise order on the usage vectors.
We then consider the category whose objects are posets and whose morphisms are
relations $R : A \rel B$ satisfying the contravariant-covariant law
$\forall x,x',y,y'.~x' \leq x \to y \leq y' \to xRy \to x'Ry'$.
This category can be given the usual monoidal product of relations, which is
the pointwise product of posets on objects.
Then, we will always expect zero and addition to together form a cocommutative
comonoid in this category.
With this structure, we can get the following equivalences and functions.

@@

I do not use algebraic properties of multiplication in conjunction with
manipulation of bunched connectives in this thesis, but we could expect
scalar multiplication
to add a comodule structure over cosemiring $R$ to the cocommutative comonoid
given by zero and addition.

>>>>

The first two connectives are those we have already seen for intuitionistic
systems --- $\dot1$ and $\dottimes$.
The prototypical rules that use $\dot1$ and $\dottimes$ are the introduction
rules for $\top$ and $\with$, respectively.
The $\dot1$ connective encodes the absence of premises, while $\dottimes$
encodes the space between premises sharing a context.
As for implication, I temporarily avoid giving a $\dotto$ connective, instead
fusing it together with $\forallb{-}$ to produce the \emph{set} of
context-preserving functions $T \rightarrowtriangle U$.
When we interpret a typing rule as a constructor of an inductive definition,
$\rightarrowtriangle$
interprets the horizontal line, reflecting the fact that the usage annotations
we start off with in the premises are those of the conclusion, corresponding to
a general principle of resource conservation.

@@

The one other change I have made between \cref{fig:lr} and \cref{fig:lr-bunched}
is to abbreviate the conjunction of $\gamma \ni x : A$ and
$\grP \leq \langle x \rvert$ as $\grP\gamma \sqni A$, and then omit the common
context between the hypothesis and conclusion of the \TirName{Var} rule to give
${} \sqni A$.
Just as I prefer to work with a full context $\Gamma$ rather than its components
$\gamma$ and $\grP$, I prefer to work with the usage-checked variables in
$\Gamma \sqni A$, rather than ever considering non-usage-checked variables in
$\gamma \ni A$ and then taking usage-correctness as an extra property.

@@

The separating conjunction $\sep$ can be seen as a decategorified version of Day
convolution~\citep{Day70}.
It also resembles the use of ternary frames in semantics of non-distributive
logics~\citep[chapter 12]{Restall1999}.

@@

For all uses of bunched connectives in this thesis, the carrier set $A$ will
form a partial order --- for example, with contexts, the order is given by the
pointwise order on the usage vectors.
We then consider the category whose objects are posets and whose morphisms are
relations $R : A \rel B$ satisfying the contravariant-covariant law
$\forall x,x',y,y'.~x' \leq x \to y \leq y' \to xRy \to x'Ry'$.
This category can be given the usual monoidal product of relations, which is
the pointwise product of posets on objects.
Then, we will always expect zero and addition to together form a cocommutative
comonoid in this category --- i.e., we have a poset $M$ and order-respecting
relations $0 : M \rel 1$ and ${+} : M \rel M \times M$ such that:

\begin{itemize}
  \item $\forall a,b.~\plr{\exists x.~{+}(a; x, b) \wedge 0(x; {\cdot})}
    \iff a \leq b$
  \item $\forall a,b.~\plr{\exists x.~{+}(a; b, x) \wedge 0(x; {\cdot})}
    \iff a \leq b$
  \item $\forall a,b,c,d.~\plr{\exists x.~{+}(a; x, d) \wedge {+}(x; b, c)}
    \iff \plr{\exists y.~{+}(a; b, y) \wedge {+}(y; c, d)}$
  \item $\forall a,b,c.~{+}(a; b, c) \iff {+}(a; c, b)$
\end{itemize}

For example, when dealing with contexts, we let $M \coloneqq \Ann^n$, with
$0(\grR; {\cdot}) \coloneqq \plr{\grR \leq \gr0}$ and
${+}(\grR; \grP, \grQ) \coloneqq \plr{\grR \leq \grP + \grQ}$, and where
the required properties follow from the commutative monoid properties of vectors
over a commutative monoid (the additive monoid of the semiring $\Ann$).
With this structure, we can get the following equivalences and functions.

@@

I do not use algebraic properties of multiplication in conjunction with
manipulation of bunched connectives in this thesis, but we could expect
scalar multiplication
to add a comodule structure over the cocommutative comonoid
given by zero and addition.
I would then expect equivalences $\gr r \cdot I^* \leftrightarrowtriangle I^*$
and $\gr r \cdot \plr{T \sep U} \leftrightarrowtriangle
\gr r \cdot T \sep \gr r \cdot U$, which both at least hold in the case of
vectors over a semiring.

### 2024-03-05 tex/lr.tex
[ correction ] first pass over chapter 4

I will not cover any operational semantics or equational theory of \name{} in
this thesis.
I will discuss a denotational semantics inspired by that of
\citet{AbelBernardy2020} in \cref{sec:den-sem}.

@@

>>>>

I will not cover any operational semantics or equational theory of \name{} in
this thesis, though the substitution principle and notion of environment I
develop in \cref{sec:ren-sub-lr} make it easy to define any desired operational
semantics.
I will discuss a denotational semantics inspired by that of
\citet{AbelBernardy2020} in \cref{sec:den-sem}.

@@

One could imagine a system in which all $\leq$-constraints in \cref{fig:lr} are
replaced by equality constraints, and the \TirName{Subuse} rule is turned into a
non-syntax-directed defining rule of the system.
Indeed, a similar system is presented by \citet{Granule18} among others.
I prefer \TirName{Subuse} to be admissible in this thesis because it means that
all algebra, including inequalities, happens in the same place.
Having all of the algebra happening in the same place sets us up for
\cref{sec:lnd}, in which I pick out some of the common patterns in the rules of
\cref{fig:lr} and partially abstract away from the algebraic details.
Using the abstractions of \cref{sec:lnd}, there is no natural place for a
\TirName{Subuse} rule, though one could imagine adapting the operators there to
introduce tightening of usage constraints.
In terms of \cref{sec:framework}, where I consider $\name$-like syntaxes in
general, one would expect to add a \TirName{Subuse} rule, similarly to the
\TirName{Var} rule present in all syntaxes, if subusaging were not already
admissible.

### 2024-03-05 tex/rec.tex
[ correction ] first pass over chapter 4

We can also give an alternative elimination rule for $\otimes$-products, as
found in Granule~\citep{Granule18,HMWO21} and the work of
\citet{AbelBernardy2020} and \citet{reed10distance}.
This modified rule allows us to, for example, have a variable
$\gr rx : A \otimes B$ for any $\gr r$, and pattern-match it into
$\gr ry : A, \gr rz : B$.
With the standard $\name$ rule from \cref{fig:lr}, the newly bound variables are
always given annotation $\gr 1$, and we would only be able to do the match in
the first place if $\gr r$ were coercible to (i.e.\ less than or equal to)
$\gr1$.
The alternative rule below is not compatible with linear logic, because it
allows us to derive
$\oc\gr r\plr{A \otimes B} \multimap \oc\gr rA \otimes \oc\gr r B$
parametric in $\gr r$, $A$, and $B$.

@@

The following additions to \name{} support a broad class of
inductive types.
I define strictly positive functors syntactically, with the only notable
restrictions being not being allowed to use the type variable $X$ in the domain
of a function type and within a $\oc$-type.
I then add least fixed points of such strictly positive functors to the syntax
of types.

@@

@@

  \begin{align*}
    \TirName{nil}\left[
    \begin{prooftree}
      \hypo{\grR \leq \gr0}
      \infer1[$I$-I]{\grR\gamma \vdash I}
      \infer1[$\oplus$-I$_0$]%
      {\grR\gamma \vdash I \oplus \plr{A \otimes \mathrm{List}_A}}
      \infer1[$\mu$-I]{\grR\gamma \vdash \mathrm{List}_A}
    \end{prooftree}
    \right.
    &&
    \TirName{cons}\left[
    \begin{prooftree}
      \hypo{\grR \leq \grP + \grQ}
      \hypo{\grP\gamma \vdash A}
      \hypo{\grQ\gamma \vdash \mathrm{List}_A}
      \infer3[$\otimes$-I]{\grR\gamma \vdash A \otimes \mathrm{List}_A}
      \infer1[$\oplus$-I$_1$]%
      {\grR\gamma \vdash I \oplus \plr{A \otimes \mathrm{List}_A}}
      \infer1[$\mu$-I]{\grR\gamma \vdash \mathrm{List}_A}
    \end{prooftree}
    \right.
  \end{align*}
  \begin{displaymath}
    \TirName{foldr}\left[
    \begin{prooftree}
      \hypo{\grP\gamma \vdash \mathrm{List}_A}
      \infer0[Var]{\gr0\gamma, \gr1\plr{I \oplus \plr{A \otimes C}}
        \vdash I \oplus \plr{A \otimes C}}
      \hypo{\nabla^n}
      \hypo{\nabla^c}
      \infer3[$\oplus$-E]{\grQ\gamma, \gr1\plr{I \oplus \plr{A \otimes C}}
        \vdash C}
      \infer2[$\mu$-E]{\grR\gamma \vdash C}
    \end{prooftree}
    \right.
  \end{displaymath}
  \begin{align*}
    \textrm{where }\nabla^n &\coloneqq
    \begin{prooftree}
      \infer0[Var]{\gr0\gamma, \gr1I \vdash I}
      \hypo{\grQ\gamma \vdash C}
      \infer1[Wk]{\grQ\gamma, \gr0I \vdash C}
      \infer2[$I$-E]{\grQ\gamma, \gr1I \vdash C}
      \infer1[Wk]{\grQ\gamma, \gr0\plr{I \oplus \plr{A \otimes C}}, \gr1I
        \vdash C}
    \end{prooftree}
    \\\\
    \textrm{and }\nabla^c &\coloneqq
    \begin{prooftree}
      \infer0[Var]{\gr0\gamma, \gr1\plr{A \otimes C} \vdash A \otimes C}
      \hypo{\grQ\gamma, \gr1A, \gr1C \vdash C}
      \infer1[Wk]{\grQ\gamma, \gr0\plr{A \otimes C}, \gr1A, \gr1C \vdash C}
      \infer2[$\otimes$-E]{\grQ\gamma, \gr1\plr{A \otimes C} \vdash C}
      \infer1[Wk]%
      {\grQ\gamma, \gr0\plr{I \oplus \plr{A \otimes C}}, \gr1\plr{A \otimes C}
        \vdash C}
    \end{prooftree}
  \end{align*}
\end{example}

@@

\ExecuteMetaData[\Bunchedtex]{Box}

@@

\begin{example}
  Let us define \emph{$\with$-lists} as follows.
  \[
    \mathrm{WithList}_A \coloneqq \mu X.~\top \oplus \plr{A \with X}
  \]
  Unlike ordinary lists (which we may call \emph{$\otimes$-lists}),
  $\with$-lists support a \emph{lookup} operation with the type below, where
  $\mathbb N \coloneqq \mu X.~I \oplus X$.
  \begin{align*}
    \mathit{lookup} :
    \mathrm{WithList}_A \multimap \mathbb N \multimap \top \oplus A
  \end{align*}
  The $\name$ term implementing $\mathit{lookup}$ in terms of iterators
  (\TirName{$\mu$-E}) is complex and tedious, so I do not show it here.
  Instead, I provide the following clausal specification, where
  $\mathrm{nil}\plr{-}$ and $\mathrm{cons}\plr{-}$ are the constructors of
  $\with$-lists, $\mathrm{zero}\plr{}$ and $\mathrm{succ}\plr{}$ are the
  constructors of $\mathbb N$, $\mathrm{inl}\plr{-}$ and $\mathrm{inr}\plr{-}$
  are the constructors of $\oplus$, $\langle\rangle$ constructs the
  inhabitant of $\top$, and postfix $.\pi_l$ and $.\pi_r$ are the projections
  out of a $\with$-pair.
  %\begin{align*}
  %  \mathit{lookup}\,\mathrm{nil}\plr{t}\,i
  %  &\coloneqq \mathrm{inl}\plr{\langle\rangle} \\
  %  \mathit{lookup}\,\mathrm{cons}\plr{\mathit{xxs}}\,\mathrm{zero}\plr{}
  %  &\coloneqq \mathrm{inr}\plr{\mathit{xxs}.\pi_l} \\
  %  \mathit{lookup}\,\mathrm{cons}\plr{\mathit{xxs}}\,\mathrm{succ}\plr{i}
  %  &\coloneqq \mathit{lookup}\,\plr{\mathit{xxs}.\pi_r}\,i
  %\end{align*}

  %\noindent{}\hspace{-0.75ex}
  \[
    \begin{pmboxed}
      \defaultcolumn{l}
      \> \mathit{lookup} \> \mathrm{nil}\plr{t} \> i
      \> \coloneqq \> \mathrm{inl}\plr{\langle\rangle} \\
      \> \mathit{lookup} \> \mathrm{cons}\plr{\mathit{xxs}}
      \> \mathrm{zero}\plr{} \> \coloneqq
      \> \mathrm{inr}\plr{\mathit{xxs}.\pi_l} \\
      \> \mathit{lookup} \> \mathrm{cons}\plr{\mathit{xxs}}
      \> \mathrm{succ}\plr{i} \> \coloneqq
      \> \mathit{lookup}\,\plr{\mathit{xxs}.\pi_r}\,i \\
    \end{pmboxed}
  \]
  %\begin{verbatim}
  %  lookup := \xs. wfoldr(xs;
  %    List I -o T + A;
  %    t. \i. foldr(i; T -o T + A; \t'. inl(t'); u, r. let () := u in r) t;
  %    xr. \i. let (i', r) :=
  %      foldr(i;
  %        A & (T + A) -o List I * (T + A);
  %        \xr'. ([], prl(xr'));
  %        u, ir. \xr'.
  %          let () := u in
  %          let (i', r) := ir in
  %          (() :: i', prr(xr') prl(ir)))
  %      xr in
  %      foldr(i'; T + A; r; u, r'. let () := u in r'))
  %\end{verbatim}
  %\begin{align*}
  %  \mathit{lookup} &:
  %  \mathrm{WithList}_A \multimap \mathrm{List}_I \multimap \top \oplus A
  %  \\
  %                  &\coloneqq \lambda\mathit{xs}.~\mathrm{rec}\plr{xs; %
  %                    \mathrm{List}_I \multimap \top \oplus A; %
  %                    t.~\lambda i.~\mathrm{inl}\plr{%
  %                    \mathrm{foldr}\plr{i, \top \multimap \top; \lambda t'.~t'; %
  %                    u, r.~\textrm{let }{*} \coloneqq u\textrm{ in }r}\,t}; %
  %                    \mathit{xr}.~\lambda i.~%
  %                    \mathrm{foldr}\plr{i; %
  %                    A \with \plr{\top \oplus A} \multimap \top \oplus A; %
  %                    \lambda
  %                    }\,\mathit{xr}}
  %\end{align*}
\end{example}

>>>>

We can also give an alternative elimination rule for $\otimes$-products, as
found in Granule~\citep{Granule18,HMWO21} and the work of
\citet{AbelBernardy2020} and \citet{reed10distance}.
This modified rule allows us to, for example, have a variable
$\gr rx : A \otimes B$ for any $\gr r$, and pattern-match it into
$\gr ry : A, \gr rz : B$.
With the standard $\name$ rule from \cref{fig:lr}, the newly bound variables are
always given annotation $\gr 1$, and we would only be able to do the match in
the first place if $\gr r$ were coercible to (i.e.\ less than or equal to)
$\gr1$.
The alternative rule below is not compatible with linear logic, because it
allows us to derive
$\oc\gr r\plr{A \otimes B} \multimap \oc\gr rA \otimes \oc\gr r B$
parametric in $\gr r$, $A$, and $B$.
In particular, with the linearity semiring of \cref{def:lin-semiring}, we get
$\oc\gr\omega\plr{A \otimes B} \multimap \oc\gr\omega A \otimes \oc\gr\omega B$,
corresponding to the ordinary linear logic formula
$\oc\plr{A \otimes B} \multimap \oc A \otimes \oc B$, which is not provable.

@@

The following additions to \name{} support a broad class of
inductive types.
I define strictly positive functors syntactically, with the only notable
restrictions relative to the syntax of types with a type variable $X$
being not being allowed to use $X$ in the domain of a function type or within a
$\oc$-type.
I then add least fixed points of such strictly positive functors to the syntax
of types.

@@

One could also consider supporting mutually inductive types, but I do not in
this thesis for the sake of simplicity.

@@

  \begin{align*}
    \TirName{nil}\left[
    \begin{prooftree}
      \hypo{\grR \leq \gr0}
      \infer1[$I$-I]{\grR\gamma \vdash I}
      \infer1[$\oplus$-I$_0$]%
      {\grR\gamma \vdash I \oplus \plr{A \otimes \mathrm{List}_A}}
      \infer1[$\mu$-I]{\grR\gamma \vdash \mathrm{List}_A}
    \end{prooftree}
    \right.
    &&
    \TirName{cons}\left[
    \begin{prooftree}
      \hypo{\grR \leq \grP + \grQ}
      \hypo{\grP\gamma \vdash A}
      \hypo{\grQ\gamma \vdash \mathrm{List}_A}
      \infer3[$\otimes$-I]{\grR\gamma \vdash A \otimes \mathrm{List}_A}
      \infer1[$\oplus$-I$_1$]%
      {\grR\gamma \vdash I \oplus \plr{A \otimes \mathrm{List}_A}}
      \infer1[$\mu$-I]{\grR\gamma \vdash \mathrm{List}_A}
    \end{prooftree}
    \right.
  \end{align*}
  \begin{displaymath}
    \TirName{foldr}\left[
    \begin{prooftree}
      \hypo{\grP\gamma \vdash \mathrm{List}_A}
      \infer0[Var]{\gr0\gamma, \gr1\plr{I \oplus \plr{A \otimes C}}
        \vdash I \oplus \plr{A \otimes C}}
      \hypo{\nabla^n}
      \hypo{\nabla^c}
      \infer3[$\oplus$-E]{\grQ\gamma, \gr1\plr{I \oplus \plr{A \otimes C}}
        \vdash C}
      \infer2[$\mu$-E]{\grR\gamma \vdash C}
    \end{prooftree}
    \right.
  \end{displaymath}
  \begin{align*}
    \textrm{where }\nabla^n &\coloneqq
    \begin{prooftree}
      \infer0[Var]{\gr0\gamma, \gr1I \vdash I}
      \hypo{\grQ\gamma \vdash C}
      \infer1[Wk]{\grQ\gamma, \gr0I \vdash C}
      \infer2[$I$-E]{\grQ\gamma, \gr1I \vdash C}
      \infer1[Wk]{\grQ\gamma, \gr0\plr{I \oplus \plr{A \otimes C}}, \gr1I
        \vdash C}
    \end{prooftree}
    \\\\
    \textrm{and }\nabla^c &\coloneqq
    \begin{prooftree}
      \infer0[Var]{\gr0\gamma, \gr1\plr{A \otimes C} \vdash A \otimes C}
      \hypo{\grQ\gamma, \gr1A, \gr1C \vdash C}
      \infer1[Wk]{\grQ\gamma, \gr0\plr{A \otimes C}, \gr1A, \gr1C \vdash C}
      \infer2[$\otimes$-E]{\grQ\gamma, \gr1\plr{A \otimes C} \vdash C}
      \infer1[Wk]%
      {\grQ\gamma, \gr0\plr{I \oplus \plr{A \otimes C}}, \gr1\plr{A \otimes C}
        \vdash C}
    \end{prooftree}
  \end{align*}

  Specially for lists, we can note that the \TirName{nil} constructor appears
  exactly once (at the end of a series of \TirName{cons}es) in any value.
  This suggests that we can handle \TirName{nil} linearly when doing recursion
  over lists --- i.e., we do not need to handle the \TirName{nil} constructor in
  a usage context $\grQ$ satisfying $\grQ \leq \gr0$ and
  $\grQ \leq \grQ + \grQ$.

  The key in treating \TirName{nil} linearly is to use \TirName{foldr} with
  endofunction return type $C \multimap C$.
  Such functions let us pass the desired \TirName{nil} value through all of the
  \TirName{cons}es into its required place.
  I give this construction rigorously in the derivation below, which describes
  how to take a list, a \TirName{nil} value $\nabla^n : \grS\gamma \vdash C$,
  and a \TirName{cons} value $\nabla^c : \grQ\gamma, \gr1A, \gr1C \vdash C$ and
  produce a result of type $C$.
  I have omitted side conditions
  $\grR \leq \grP + \grQ$, $\grQ \leq \gr0$, $\grQ \leq \grQ + \grQ$, and
  $\grT \leq \grR + \grS$ in order to save space in the figure.
  I have also omitted subderivation $\mathrm{id}$, constructing the identity
  function (noting that $\grQ \leq \gr0$, and therefore all of the additional
  variables can be discarded), and the details of a use of substitution ---
  labelled \TirName{Subst}.
  I introduce substitution in its full generality in \cref{sec:ren-sub-lr}, but
  the intuitive justification for using substitution here is that we apply the
  variable of type $C \multimap C$ to the variable of type $C$ to produce a
  value of type $C$.
  This uses up both the function and the argument, leaving just a single linear
  value.
  One could also imagine introducing a let-binding, together with weakening of
  $\gr0$-annotated variables, to make the shape of the sequent line up with that
  of $\nabla^c$.

  \begin{displaymath}
    \begin{prooftree}
      \hypo{\grP\gamma \vdash \mathrm{List}_A}
      \hypo{\mathrm{id}}
      \infer[no rule]1{\grQ\gamma \vdash C \multimap C}
      \hypo{\nabla^c}
      \infer[no rule]1{\grQ\gamma, \gr1A, \gr1C \vdash C}
      \infer1[Subst]{\grQ\gamma, \gr1A, \gr1\plr{C \multimap C}, \gr1C \vdash C}
      \infer1[$\multimap$-I]{\grQ\gamma, \gr1A, \gr1\plr{C \multimap C} \vdash C \multimap C}
      \infer3[foldr]{\grR\gamma \vdash C \multimap C}
      \hypo{\nabla^n}
      \infer[no rule]1{\grS\gamma \vdash C}
      \infer2[$\multimap$-E]{\grT\gamma \vdash C}
    \end{prooftree}
  \end{displaymath}

  Given an annotation $\gr r$ satisfying $\gr r \leq \gr0$,
  $\gr r \leq \gr r + \gr r$, and $\gr r \leq 1$ (for example, $\gr\omega$ from
  \cref{def:lin-semiring}), we can produce a function
  \[
    \mathrm{foldr}_{A,C} : \mathrm{List}_A \multimap \oc\gr rC \multimap
    \oc\gr r\plr{A \multimap C \multimap C} \multimap C
  \]
  as a simple wrapping of the \TirName{foldr} rule.
  We can get a function $\mathrm{foldr}_{A,C}'$ as follows using
  $\mathrm{foldr}_{A,C \multimap C}$, and similar reasoning as above to
  treat the \TirName{nil} case linearly.
  I use square brackets to denote the \TirName{$\oc$-I} rule, a pattern
  matching let-expression to denote the \TirName{$\oc$-E} rule, and standard
  $\lambda$-calculus notation for function introduction and elimination.
  \begin{align*}
    \mathrm{foldr}_{A,C}' &: \mathrm{List}_A \multimap C \multimap
    \oc\gr r\plr{A \multimap C \multimap C} \multimap C \\
    \mathrm{foldr}_{A,C}' &\coloneqq \lambda e.\lambda n.\lambda c'.~%
                            \textrm{let }[c] \coloneqq c'\textrm{ in }
                            \mathrm{foldr}_{A,C \multimap C}\,e\,
                            [\lambda r.~r]\,
                            [\lambda x.\lambda f.\lambda r.~c\,x\,(f\,r)]\,
                            n
  \end{align*}
\end{example}

\begin{example}
  Let us define \emph{$\with$-lists} as follows.
  \[
    \mathrm{WithList}_A \coloneqq \mu X.~\top \oplus \plr{A \with X}
  \]
  Unlike ordinary lists (which we may call \emph{$\otimes$-lists}),
  $\with$-lists support a \emph{lookup} operation with the type below, where
  $\mathbb N \coloneqq \mu X.~I \oplus X$.
  \begin{align*}
    \mathit{lookup} :
    \mathrm{WithList}_A \multimap \mathbb N \multimap \top \oplus A
  \end{align*}
  The $\name$ term implementing $\mathit{lookup}$ in terms of iterators
  (\TirName{$\mu$-E}) is complex and tedious, so I do not show it here.
  Instead, I provide the following clausal specification, where
  $\mathrm{nil}\plr{-}$ and $\mathrm{cons}\plr{-}$ are the constructors of
  $\with$-lists, $\mathrm{zero}\plr{}$ and $\mathrm{succ}\plr{}$ are the
  constructors of $\mathbb N$, $\mathrm{inl}\plr{-}$ and $\mathrm{inr}\plr{-}$
  are the constructors of $\oplus$, $\langle\rangle$ constructs the
  inhabitant of $\top$, and postfix $.\pi_l$ and $.\pi_r$ are the projections
  out of a $\with$-pair.

  \[
    \begin{pmboxed}
      \defaultcolumn{l}
      \> \mathit{lookup} \> \mathrm{nil}\plr{t} \> i
      \> \coloneqq \> \mathrm{inl}\plr{\langle\rangle} \\
      \> \mathit{lookup} \> \mathrm{cons}\plr{\mathit{xxs}}
      \> \mathrm{zero}\plr{} \> \coloneqq
      \> \mathrm{inr}\plr{\mathit{xxs}.\pi_l} \\
      \> \mathit{lookup} \> \mathrm{cons}\plr{\mathit{xxs}}
      \> \mathrm{succ}\plr{i} \> \coloneqq
      \> \mathit{lookup}\,\plr{\mathit{xxs}.\pi_r}\,i \\
    \end{pmboxed}
  \]
\end{example}

@@

\begin{samepage}
  \ExecuteMetaData[\Bunchedtex]{Box}
\end{samepage}

@@

### 2024-03-05 tex/semirings.tex
[ correction ] first pass over chapter 4

The question of defining calculi which do not semantically admit weakening and
contraction but also do not rely on variables going out of scope is directly
addressed by \citet{McBride16}.
The first technique suggested is to, instead of
removing variables from the context of certain subterms, add an annotation to
free variables saying whether or not they are to be used.
I use an annotation $\gr0$ on variables that are not to be used, and an
annotation $\gr1$ on variables that are to be used.
This convention lets us transcribe the usual $\otimes$-introduction rule
(below left) as a rule with usage annotations (below right).
In the notation on the right, I let $\Gamma = \grP\gamma$ and
$\Delta = \grQ\delta$, where $\Gamma$ is a whole context comprising a
\emph{usage context} $\grP$ and a \emph{typing context} $\gamma$.
A usage context is a list of usage annotations, so
$\grP = \gr{r_1}, \ldots, \gr{r_m}$ and a typing context is a list of types, so
$\gamma = A_1, \ldots, A_m$.
When combined, the usage context and the typing context will be of the same
length.
Explicit contexts will usually be written with usage annotations and types
interspersed, as $\gr{r_1}A_1, \ldots, \gr{r_m}A_m$.
I use $\gr r\gamma$ to abbreviate $\gr rx_1, \ldots, \gr rx_m$.

@@

The final interesting rule form to cover is found in DILL's
$\oc$-introduction rule.
DILL's $\oc$-introduction can be though of as an $\gr\omega$-ary counterpart to
$\otimes$-introduction, though with the same premise each time rather than
$\gr\omega$-many premises.
This explains why only $\gr\omega$- and
$\gr0$-annotated variables can appear in the conclusion of $\oc$-introduction,
and also justifies the choice below of multiplication (vector scaling) as the
algebraic operation controlling the $\oc$-modality.

@@

The distinction between sharing and separating conjunction of premises naturally
falls out of the posemiring approach to usage restrictions.
However, a similar distinction also appears in other approaches to linearity,
and perhaps in other substructural systems.
Indeed, in the paper that inspired the bunched connectives~\citep{RPKV20},
linearity is enforced in a similar style as in Yalla~\citep{laurent18}, using
lists which can be split.
It would be interesting to see future work on substructural systems work out the
appropriate premise connectives and define their systems from there, rather than
directly manipulating contexts.
It may also be possible to see future work profitably abstracting over the
posemiring annotations, requiring only that contexts form something like a
commutative $\mathrm{Rel}$-monoid supporting variable-binding and
variable access.

>>>>

The question of defining calculi which do not semantically admit weakening and
contraction but also do not rely on variables going out of scope is directly
addressed by \citet{McBride16}.
The first technique suggested is to, instead of
removing variables from the context of certain subterms, add an annotation to
free variables saying whether or not they are to be used.
I use an annotation $\gr0$ on variables that are not to be used, and an
annotation $\gr1$ on variables that are to be used.
This convention lets us transcribe the usual $\otimes$-introduction rule
(below left) as a rule with usage annotations (below right).
In the notation on the right, I let $\Gamma = \grP\gamma$ and
$\Delta = \grQ\delta$, where $\Gamma$ is a whole context comprising a
\emph{usage context} $\grP$ and a \emph{typing context} $\gamma$.
A usage context is a list of usage annotations, so
$\grP = \gr{r_1}, \ldots, \gr{r_m}$ and a typing context is a list of types, so
$\gamma = A_1, \ldots, A_m$.
Note that, in a single context $\Gamma$, components $\grP$ and $\gamma$ have the
same length $m$ (and the same shape when represented in Agda).
When combined, the usage context and the typing context will be of the same
length.
Explicit contexts will usually be written with usage annotations and types
interspersed, as $\gr{r_1}A_1, \ldots, \gr{r_m}A_m$.
I use $\gr r\gamma$ to abbreviate $\gr rA_1, \ldots, \gr rA_m$.

@@

The final interesting rule form to cover is found in DILL's
$\oc$-introduction rule.
DILL's $\oc$-introduction can be thought of as an $\gr\omega$-ary counterpart to
$\otimes$-introduction, though with the same premise each time rather than
$\gr\omega$-many premises.
This explains why only $\gr\omega$- and
$\gr0$-annotated variables can appear in the conclusion of $\oc$-introduction,
and also justifies the choice below of multiplication (vector scaling) as the
algebraic operation controlling the $\oc$-modality.

@@

The distinction between sharing and separating conjunction of premises naturally
falls out of the posemiring approach to usage restrictions.
However, a similar distinction also appears in other approaches to linearity,
and perhaps in other substructural systems.
Indeed, in the paper that inspired the bunched connectives~\citep{RPKV20},
linearity is enforced in a similar style as in Yalla~\citep{laurent18}, using
lists which can be split.
It would be interesting to see future work on substructural systems work out the
appropriate premise connectives and define their systems from there, rather than
directly manipulating contexts.
It may also be possible to see future work profitably abstracting over the
posemiring annotations, requiring only that contexts form something like a
commutative $\mathrm{Rel}$-monoid supporting variable binding and
variable access.
